---
title: "Unit 1: Introduction to R and Data Structures"
description: "Interactive tutorial on basic R syntax, data types, vectors, matrices, data frames, and factor variables."
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
library(readxl)
knitr::opts_chunk$set(echo = FALSE)
```

## Welcome

Welcome to Unit 1! This tutorial covers the fundamentals of R and its core data structures, including vectors, matrices, arrays, lists, factors, and data frames.

---

## Exercise 1: Comments and Timestamps

Documenting your code and recording execution times are critical for reproducible research.

### Practice Problem
1. Create a script header comment.
2. Print the current date and time using `Sys.time()`.
3. Print a custom formatted timestamp: `"Execution Timestamp: [Formatted Date & Time]"` using `cat()` and `format()`.

```{r ex1, exercise=TRUE}
# 1. Add a comment describing this script


# 2. Print current date and time using Sys.time()


# 3. Print a formatted execution timestamp


```

```{r ex1-hint-1}
# Hint 1 (Concept): Use Sys.time() to retrieve the current system time, and format() with cat() to format and display custom text.
```

```{r ex1-hint-2}
# Hint 2 (Code Skeleton):
# 1. Comment
# 2. Sys.time()
# 3. cat("Execution Timestamp:", format(Sys.time(), "___"), "\n")
```

```{r ex1-hint-3}
# Hint 3 (Detailed Guide): Use format(Sys.time(), "%Y-%m-%d %H:%M:%S") inside cat() to print an ISO-standard formatted timestamp.
```

```{r ex1-solution}
# Unit 1: Comments and Timestamps
# Author: Student

# 2. Print current date and time
Sys.time()

# 3. Print formatted timestamp
cat("Execution Timestamp:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```

```{r ex1-check}
grade_this_code()
```

---

## Exercise 2: Vector Creation and Arithmetic

Vectors are 1-dimensional arrays holding elements of the same data type.

### Practice Problem
1. Create a numeric vector `v1` containing integers from `5` to `15` using the `:` operator.
2. Create a numeric vector `v2` containing a sequence from `10` to `30` in steps of `2` using `seq()`.
3. Add the first 6 elements of `v1` and `v2` together.

```{r ex2, exercise=TRUE}
# 1. Create v1


# 2. Create v2


# 3. Add first 6 elements of v1 and v2


```

```{r ex2-hint-1}
# Hint 1 (Concept): Use the colon operator (:) for consecutive integers, seq() with from, to, and by arguments for sequences, and [] for indexing.
```

```{r ex2-hint-2}
# Hint 2 (Code Skeleton):
v1 <- 5:15
v2 <- seq(from = 10, to = 30, by = 2)
v1[1:6] + v2[1:6]
```

```{r ex2-hint-3}
# Hint 3 (Detailed Guide): Vector addition in R is element-wise. Subsetting both vectors with [1:6] ensures you add matching 6 elements.
```

```{r ex2-solution}
# 1. Create v1
v1 <- 5:15

# 2. Create v2
v2 <- seq(from = 10, to = 30, by = 2)

# 3. Add first 6 elements of v1 and v2
v1[1:6] + v2[1:6]
```

```{r ex2-check}
grade_this_code()
```

---

## Exercise 3: Vector Indexing by Position

You can subset vectors using positive indices, negative indices, or index vectors.

### Practice Problem
Using the preloaded vector `temp <- c(36.5, 37.2, 38.9, 39.1, 36.8)`:

1. Extract the 3rd element.
2. Extract the last element using `length()`.
3. Extract all elements *except* the 2nd one (using a negative index).

```{r ex3, exercise=TRUE}
temp <- c(36.5, 37.2, 38.9, 39.1, 36.8)

# 1. Extract 3rd element


# 2. Extract last element


# 3. Extract all except the 2nd element


```

```{r ex3-hint-1}
# Hint 1 (Concept): Use square brackets [] to extract elements by position. To exclude an element, use a negative index like [-2].
```

```{r ex3-hint-2}
# Hint 2 (Code Skeleton):
temp[3]
temp[length(temp)]
temp[-2]
```

```{r ex3-hint-3}
# Hint 3 (Detailed Guide): length(temp) dynamically finds the index of the last element, regardless of how long temp is.
```

```{r ex3-solution}
temp <- c(36.5, 37.2, 38.9, 39.1, 36.8)

# 1. Extract 3rd element
temp[3]

# 2. Extract last element
temp[length(temp)]

# 3. Extract all except 2nd element
temp[-2]
```

```{r ex3-check}
grade_this_code()
```

---

## Exercise 4: Vector Indexing by Logical Conditions

Logical subsetting filters vectors by keeping only elements that satisfy a condition.

### Practice Problem
Using the preloaded vector `scores <- c(45, 82, 91, 58, 73, 64, 88)`:

1. Find which scores are greater than or equal to `70` (returns a logical vector).
2. Subset `scores` to extract the actual values that are greater than or equal to `70`.

```{r ex4, exercise=TRUE}
scores <- c(45, 82, 91, 58, 73, 64, 88)

# 1. Find logical vector for scores >= 70


# 2. Extract scores >= 70


```

```{r ex4-hint-1}
# Hint 1 (Concept): Apply logical operator >= directly to a vector to get a boolean vector. Place that condition inside [] to subset the values.
```

```{r ex4-hint-2}
# Hint 2 (Code Skeleton):
scores >= 70
scores[scores >= 70]
```

```{r ex4-hint-3}
# Hint 3 (Detailed Guide): Logical indexing filters the vector keeping only TRUE elements. `scores >= 70` evaluates to logical TRUE/FALSE values.
```

```{r ex4-solution}
scores <- c(45, 82, 91, 58, 73, 64, 88)

# 1. Logical vector for scores >= 70
scores >= 70

# 2. Extract scores >= 70
scores[scores >= 70]
```

```{r ex4-check}
grade_this_code()
```

---

## Exercise 5: Mathematical and Rounding Functions

R provides built-in functions for mathematical calculations and rounding.

### Practice Problem
1. Find the absolute values of the vector `x <- c(-5.6, 2.3, -9.1)`.
2. Compute the natural logarithm (`log()`) of `100`, and round the result to `2` decimal places.
3. Compute the square root of `50`, and round it down to the nearest integer using `floor()`.

```{r ex5, exercise=TRUE}
# 1. Absolute values of x


# 2. Log of 100 rounded to 2 decimal places


# 3. Square root of 50 rounded down


```

```{r ex5-hint-1}
# Hint 1 (Concept): Use abs() for absolute values, log() for natural logarithm, round() for rounding to decimals, and floor() to round down.
```

```{r ex5-hint-2}
# Hint 2 (Code Skeleton):
abs(x)
round(log(100), digits = 2)
floor(sqrt(50))
```

```{r ex5-hint-3}
# Hint 3 (Detailed Guide): log() in R defaults to natural logarithm (base e). Use digits = 2 in round() to keep two decimal places.
```

```{r ex5-solution}
x <- c(-5.6, 2.3, -9.1)

# 1. Absolute values
abs(x)

# 2. Log of 100 rounded to 2 decimals
round(log(100), 2)

# 3. Square root of 50 rounded down
floor(sqrt(50))
```

```{r ex5-check}
grade_this_code()
```

---

## Exercise 6: Matrix Creation

A matrix is a 2-dimensional vector with rows and columns.

### Practice Problem
1. Create a 3x3 matrix named `my_mat` containing values `1` to `9`, filled **by row**.
2. Assign row names `"Row1", "Row2", "Row3"` and column names `"Col1", "Col2", "Col3"`.

```{r ex6, exercise=TRUE}
# 1. Create a 3x3 matrix filled by row


# 2. Assign rownames and colnames


```

```{r ex6-hint-1}
# Hint 1 (Concept): Use matrix() with nrow, ncol, and byrow = TRUE. Use rownames() and colnames() to assign dimension names.
```

```{r ex6-hint-2}
# Hint 2 (Code Skeleton):
my_mat <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)
rownames(my_mat) <- c("Row1", "Row2", "Row3")
colnames(my_mat) <- c("Col1", "Col2", "Col3")
```

```{r ex6-hint-3}
# Hint 3 (Detailed Guide): Setting `byrow = TRUE` fills the matrix row-by-row (1,2,3 in row 1; 4,5,6 in row 2; 7,8,9 in row 3).
```

```{r ex6-solution}
# 1. Create 3x3 matrix filled by row
my_mat <- matrix(1:9, nrow = 3, ncol = 3, byrow = TRUE)

# 2. Assign rownames and colnames
rownames(my_mat) <- c("Row1", "Row2", "Row3")
colnames(my_mat) <- c("Col1", "Col2", "Col3")
my_mat
```

```{r ex6-check}
grade_this_code()
```

---

## Exercise 7: Matrix Subsetting

You can subset a matrix using `my_mat[row, column]`.

### Practice Problem
Using the preloaded matrix `mat`:

1. Extract the entire 2nd row.
2. Extract the entire 3rd column.
3. Extract the element at the intersection of the 2nd row and 3rd column.

```{r ex7, exercise=TRUE}
mat <- matrix(10:18, nrow = 3, byrow = TRUE)

# 1. Extract 2nd row


# 2. Extract 3rd column


# 3. Extract element at row 2, col 3


```

```{r ex7-hint-1}
# Hint 1 (Concept): Subset a matrix using mat[row, col]. Leaving row or column blank extracts the entire row or column.
```

```{r ex7-hint-2}
# Hint 2 (Code Skeleton):
mat[2, ]
mat[, 3]
mat[2, 3]
```

```{r ex7-hint-3}
# Hint 3 (Detailed Guide): `mat[2, ]` selects row 2 and all columns. `mat[, 3]` selects all rows and column 3.
```

```{r ex7-solution}
mat <- matrix(10:18, nrow = 3, byrow = TRUE)

# 1. Extract 2nd row
mat[2, ]

# 2. Extract 3rd column
mat[, 3]

# 3. Extract element at row 2, col 3
mat[2, 3]
```

```{r ex7-check}
grade_this_code()
```

---

## Exercise 8: Matrix Arithmetic & Row/Column Statistics

R provides convenient functions for summarizing matrix rows and columns.

### Practice Problem
Using the preloaded matrix `mat`:

1. Calculate the sum of each row using `rowSums()`.
2. Calculate the mean of each column using `colMeans()`.

```{r ex8, exercise=TRUE}
mat <- matrix(1:12, nrow = 3)

# 1. Sum of each row


# 2. Mean of each column


```

```{r ex8-hint-1}
# Hint 1 (Concept): Use built-in matrix summary functions rowSums() for row totals and colMeans() for column averages.
```

```{r ex8-hint-2}
# Hint 2 (Code Skeleton):
rowSums(mat)
colMeans(mat)
```

```{r ex8-hint-3}
# Hint 3 (Detailed Guide): rowSums(mat) operates across columns for each row, returning a numeric vector of row sums.
```

```{r ex8-solution}
mat <- matrix(1:12, nrow = 3)

# 1. Sum of each row
rowSums(mat)

# 2. Mean of each column
colMeans(mat)
```

```{r ex8-check}
grade_this_code()
```

---

## Exercise 9: Arrays

Arrays generalize matrices to three or more dimensions.

### Practice Problem
1. Create a 3-dimensional array named `my_arr` containing numbers `1` to `12`.
2. Give it dimensions of `2` rows, `3` columns, and `2` layers (slices) using `array()`.

```{r ex9, exercise=TRUE}
# 1. Create a 3D array of size 2x3x2


```

```{r ex9-hint-1}
# Hint 1 (Concept): Use array() specifying the data vector and a dim argument as a numeric vector c(rows, cols, layers).
```

```{r ex9-hint-2}
# Hint 2 (Code Skeleton):
my_arr <- array(1:12, dim = c(2, 3, 2))
```

```{r ex9-hint-3}
# Hint 3 (Detailed Guide): `dim = c(2, 3, 2)` creates an array with 2 rows, 3 columns, and 2 slices/layers.
```

```{r ex9-solution}
# 1. Create a 3D array of size 2x3x2
my_arr <- array(1:12, dim = c(2, 3, 2))
my_arr
```

```{r ex9-check}
grade_this_code()
```

---

## Exercise 10: List Creation

Lists are ordered containers that can hold elements of different modes and lengths.

### Practice Problem
1. Create a list named `patient_record` containing:
   - A numeric scalar named `id` with value `402`.
   - A character vector named `symptoms` with values `"cough"`, `"fever"`.
   - A logical scalar named `admitted` with value `TRUE`.

```{r ex10, exercise=TRUE}
# 1. Create patient_record list


```

```{r ex10-hint-1}
# Hint 1 (Concept): Use list() with named arguments to store heterogeneous data types (numeric, character vector, logical).
```

```{r ex10-hint-2}
# Hint 2 (Code Skeleton):
patient_record <- list(
  id = 402,
  symptoms = c("cough", "fever"),
  admitted = TRUE
)
```

```{r ex10-hint-3}
# Hint 3 (Detailed Guide): Each element in a list can be a different data type and length. Use named tags like `id = `, `symptoms = `.
```

```{r ex10-solution}
# 1. Create patient_record list
patient_record <- list(
  id = 402,
  symptoms = c("cough", "fever"),
  admitted = TRUE
)
patient_record
```

```{r ex10-check}
grade_this_code()
```

---

## Exercise 11: List Indexing

You can extract items from a list using `[ ]`, `[[ ]]`, or `$`.

### Practice Problem
Using the preloaded list `record`:

1. Extract the `symptoms` vector as a simple character vector (using `[[ ]]` or `$`).
2. Extract a sublist containing *only* the `id` element (using `[ ]`).

```{r ex11, exercise=TRUE}
record <- list(id = 402, symptoms = c("cough", "fever"), admitted = TRUE)

# 1. Extract symptoms vector


# 2. Extract id sublist


```

```{r ex11-hint-1}
# Hint 1 (Concept): Use $ or [[ ]] to extract list contents directly as a vector. Use [ ] to extract a sublist container.
```

```{r ex11-hint-2}
# Hint 2 (Code Skeleton):
record$symptoms   # or record[["symptoms"]]
record["id"]
```

```{r ex11-hint-3}
# Hint 3 (Detailed Guide): `record$symptoms` extracts the character vector. `record["id"]` returns a sublist containing `id`.
```

```{r ex11-solution}
record <- list(id = 402, symptoms = c("cough", "fever"), admitted = TRUE)

# 1. Extract symptoms vector
record$symptoms

# 2. Extract id sublist
record["id"]
```

```{r ex11-check}
grade_this_code()
```

---

## Exercise 12: Factors

Factors represent categorical variables with predefined levels.

### Practice Problem
Consider the character vector `responses <- c("Agree", "Disagree", "Neutral", "Agree")`.

1. Convert `responses` into an ordered factor named `ordered_responses`.
2. Explicitly define the levels order as `"Disagree"`, `"Neutral"`, `"Agree"`.

```{r ex12, exercise=TRUE}
responses <- c("Agree", "Disagree", "Neutral", "Agree")

# 1. Create ordered_responses factor


```

```{r ex12-hint-1}
# Hint 1 (Concept): Convert character vectors to ordered factors using factor() with levels = c(...) and ordered = TRUE.
```

```{r ex12-hint-2}
# Hint 2 (Code Skeleton):
ordered_responses <- factor(responses, levels = c("Disagree", "Neutral", "Agree"), ordered = TRUE)
```

```{r ex12-hint-3}
# Hint 3 (Detailed Guide): `levels` defines the explicit order from lowest to highest. Setting `ordered = TRUE` establishes ordinal rank.
```

```{r ex12-solution}
responses <- c("Agree", "Disagree", "Neutral", "Agree")

# 1. Create ordered_responses factor
ordered_responses <- factor(responses, levels = c("Disagree", "Neutral", "Agree"), ordered = TRUE)
ordered_responses
```

```{r ex12-check}
grade_this_code()
```

---

## Exercise 13: Data Frames and Data Imports

A data frame is a list of equal-length vectors, mimicking a dataset table.

### Practice Problem
1. Load the pre-packaged dataset `student_scores.csv` from the path `data/student_scores.csv` using `read.csv()`.
2. Store it in a variable named `df`.
3. Print the number of rows of `df` using `nrow()`, and view the first 3 lines of `df` using `head()`.

```{r ex13, exercise=TRUE}
# 1. Read student_scores.csv into df


# 2. Print row count and first 3 lines


```

```{r ex13-hint-1}
# Hint 1 (Concept): Use read.csv() to load CSV data into a data frame. Use nrow() for row count and head(df, n = 3) for top rows.
```

```{r ex13-hint-2}
# Hint 2 (Code Skeleton):
df <- read.csv("data/student_scores.csv")
nrow(df)
head(df, n = 3)
```

```{r ex13-hint-3}
# Hint 3 (Detailed Guide): `read.csv()` imports CSV files. `head(df, 3)` displays the first 3 rows.
```

```{r ex13-solution}
# 1. Read student_scores.csv into df
df <- read.csv("data/student_scores.csv")

# 2. Print row count and first 3 lines
nrow(df)
head(df, n = 3)
```

```{r ex13-check}
grade_this_code()
```

---

## Concept Check Quiz

Test your understanding of Unit 1!

```{r quiz}
quiz(
  caption = "Unit 1 Knowledge Check (10 Questions)",
  question("What is the main difference between a matrix and a data frame in R?",
    answer("A matrix can only hold a single data type, while a data frame can hold multiple types in different columns.", correct = TRUE),
    answer("A matrix is 3-dimensional, while a data frame is 2-dimensional.", message = "No, both matrices and data frames are 2-dimensional."),
    answer("A data frame can only hold numbers, while a matrix holds text.", message = "Incorrect. Matrices can hold text or numbers, but all elements must be of the same single type."),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question_text(
    "What R function is used to load data from a comma-separated CSV file into a data frame?",
    answer("read.csv", correct = TRUE), answer("read.csv()", correct = TRUE),
    allow_retry = TRUE, trim = TRUE, options = list(placeholder = "e.g., function_name")
  ),
  question_numeric(
    "How many dimensions does a standard R matrix have?",
    answer(2, correct = TRUE), allow_retry = TRUE,
    correct = "Correct! A matrix is a 2-dimensional data structure (rows and columns)."
  ),
  question("If a vector contains c(10, 'apple', TRUE), what data type will R coerce all elements to?",
    answer("Character", correct = TRUE),
    answer("Numeric", message = "No, strings cannot be implicitly converted to numbers."),
    answer("Logical", message = "No, text strings force implicit coercion to character."),
    allow_retry = TRUE
  ),
  question("Which operator extracts list elements as their actual data type (instead of a sublist container)?",
    answer("`[`", message = "No, single square brackets [ ] return a sublist container."),
    answer("`[[`", correct = TRUE),
    answer("`$`", correct = TRUE),
    allow_retry = TRUE, type = "learnr_checkbox"
  ),
  question("Which function provides a compact structural display of an R object, showing column names, types, and first few values?",
    answer("`str()`", correct = TRUE),
    answer("`summary()`", message = "No, summary() computes summary statistics."),
    answer("`head()`", message = "No, head() prints the top rows of a dataset."),
    allow_retry = TRUE
  ),
  question_numeric(
    "If an array is created with dim = c(3, 4, 2), how many total numeric elements does it contain?",
    answer(24, correct = TRUE), allow_retry = TRUE,
    correct = "Correct! 3 * 4 * 2 = 24 elements."
  ),
  question("Which R syntax correctly selects the element in row 3, column 2 of a matrix M?",
    answer("`M[3, 2]`", correct = TRUE),
    answer("`M[2, 3]`", message = "No, in R indexing is M[row, column]."),
    answer("`M[[3]][2]`", message = "No, double brackets are used for list indexing."),
    allow_retry = TRUE
  ),
  question("What argument in factor() is set to TRUE to create an ordinal (ordered) factor?",
    answer("`ordered`", correct = TRUE),
    answer("`levels`", message = "No, levels specifies the category order."),
    answer("`sorted`", message = "No, the argument name is ordered = TRUE."),
    allow_retry = TRUE
  ),
  question("Which element-wise logical operator performs element-by-element logical AND across vectors?",
    answer("`&`", correct = TRUE),
    answer("`&&`", message = "No, && evaluates only the first element of a vector."),
    allow_retry = TRUE
  )
)
```
