---
title: "Unit 3: Confidence Interval Estimation and Resampling Methods"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
library(boot)
knitr::opts_chunk$set(echo = FALSE)
income <- read.csv("data/survey_income.csv")
```

## Welcome

Welcome to Unit 3! This tutorial covers statistical estimation, manual and R-based confidence intervals for means and proportions, and resampling methods (specifically the bootstrap algorithm for estimating standard errors and various confidence intervals).

---


---

## Interactive Explorer: Sample Size & Confidence Interval Width

Adjust the sample size $n$ and confidence level to see how the width of the confidence interval reacts live!

```{r ci-demo-ui}
sliderInput("n_sim", "Sample Size (n):", min = 10, max = 250, value = 30, step = 10)
selectInput("conf_level", "Confidence Level:", choices = c("90%" = 0.90, "95%" = 0.95, "99%" = 0.99), selected = 0.95)
plotOutput("ci_plot", height = "300px")
```

```{r ci-demo-server, context="server"}
output$ci_plot <- renderPlot({
  set.seed(42)
  n <- input$n_sim
  conf <- as.numeric(input$conf_level)
  
  # Simulate sample from Normal(50, 10)
  x <- rnorm(n, mean = 50, sd = 10)
  xbar <- mean(x)
  se <- sd(x) / sqrt(n)
  alpha <- 1 - conf
  t_crit <- qt(1 - alpha / 2, df = n - 1)
  
  lower <- xbar - t_crit * se
  upper <- xbar + t_crit * se
  width <- upper - lower
  
  plot(50, 1, xlim = c(40, 60), ylim = c(0.5, 1.5), type = "n", yaxt = "n",
       xlab = "Population Mean (mu)", ylab = "",
       main = paste(conf*100, "% CI: [", round(lower, 2), ",", round(upper, 2), "] | Margin of Error = +/-", round(t_crit*se, 2)))
  
  abline(v = 50, col = "red", lty = 2, lwd = 2)
  arrows(lower, 1, upper, 1, angle = 90, code = 3, length = 0.1, col = "#2980B9", lwd = 4)
  points(xbar, 1, pch = 19, col = "#1A5276", cex = 2)
  legend("topleft", legend = c("True Mean (50)", "Sample Mean", "Confidence Interval"),
         col = c("red", "#1A5276", "#2980B9"), lty = c(2, NA, 1), pch = c(NA, 19, NA), lwd = c(2, NA, 4))
})
```


## Exercise 1: Manual t-Test Calculations

Understanding the manual calculation of the $t$-statistic and its corresponding $p$-value.

### Practice Problem
Using base R mathematical functions and the standard `mtcars$mpg` vector:

1. Compute the sample size `n`, mean `xbar`, and standard deviation `s`.
2. Compute the standard error of the mean: `se <- s / sqrt(n)`.
3. Compute the $t$-statistic for testing the null hypothesis $H_0: \mu = 20$:
   $$t = \frac{\bar{x} - \mu_0}{SE}$$

4. Calculate the two-tailed $p$-value using the cumulative density function `pt()`.

```{r ex1, exercise=TRUE}
# 1. Compute n, xbar, s


# 2. Compute standard error se


# 3. Calculate t-statistic t_stat (null mu0 = 20)


# 4. Calculate two-tailed p-value p_val


```

```{r ex1-hint-1}
# Hint 1 (Concept): Calculate sample statistics (n, xbar, s), compute standard error se = s/sqrt(n), t-statistic = (xbar - mu0)/se, and use 2*(1 - pt(abs(t_stat), df = n-1)) for the p-value.
```

```{r ex1-hint-2}
# Hint 2 (Code Skeleton):
n <- length(mtcars$mpg)
xbar <- mean(mtcars$mpg)
s <- sd(mtcars$mpg)
se <- s / sqrt(n)
t_stat <- (xbar - 20) / se
p_val <- 2 * (1 - pt(abs(t_stat), df = n - 1))
```

```{r ex1-hint-3}
# Hint 3 (Detailed Guide): `pt()` gives the cumulative density function of the t-distribution. Multiply `1 - pt(abs(t_stat), df = n-1)` by 2 for a two-tailed test.
```

```{r ex1-solution}
n <- length(mtcars$mpg)
xbar <- mean(mtcars$mpg)
s <- sd(mtcars$mpg)

# 2. Compute standard error se
se <- s / sqrt(n)

# 3. Calculate t-statistic (null mu0 = 20)
t_stat <- (xbar - 20) / se

# 4. Calculate two-tailed p-value
p_val <- 2 * (1 - pt(abs(t_stat), df = n - 1))
cat("t-stat:", t_stat, "p-val:", p_val, "\n")
```

```{r ex1-check}
grade_this_code()
```

---

## Exercise 2: Using the t.test() Function

R's built-in `t.test()` computes standard parametric confidence intervals.

### Practice Problem
Using `t.test()` and the `mtcars$mpg` vector:

1. Run a one-sample $t$-test testing the null hypothesis $H_0: \mu = 20$.
2. Store the output in a variable named `result`.
3. Extract the 95% confidence interval endpoint vector from `result$conf.int`.

```{r ex2, exercise=TRUE}
# 1. Run t.test and store in result


# 2. Extract confidence interval vector


```

```{r ex2-hint-1}
# Hint 1 (Concept): Use t.test(x, mu = 20) to run the one-sample t-test. Extract the confidence interval using result$conf.int.
```

```{r ex2-hint-2}
# Hint 2 (Code Skeleton):
result <- t.test(mtcars$mpg, mu = 20)
result$conf.int
```

```{r ex2-hint-3}
# Hint 3 (Detailed Guide): `t.test()` returns a list of class 'htest'. The confidence interval is stored in the `$conf.int` attribute.
```

```{r ex2-solution}
# 1. Run t.test and store in result
result <- t.test(mtcars$mpg, mu = 20)

# 2. Extract confidence interval vector
result$conf.int
```

```{r ex2-check}
grade_this_code()
```

---

## Exercise 3: Confidence Intervals for Proportions

R's `prop.test()` calculates standard confidence intervals for binary outcomes.

### Practice Problem
Consider a clinical study where 25 out of 60 patients reported complete pain relief.

1. Run a one-sample proportions test **without** Yates' continuity correction (`correct = FALSE`).
2. Run the proportions test **with** Yates' continuity correction (default).

```{r ex3, exercise=TRUE}
# 1. Run prop.test without correction


# 2. Run prop.test with correction


```

```{r ex3-hint-1}
# Hint: Use prop.test() specifying the number of successes, total sample size, and correct = FALSE to disable Yates' correction.
```

```{r ex3-solution}
  # solution/hint content purged from public repository history

```

```{r ex3-check}
grade_this_code()
```

---

## Exercise 4: Defining a Bootstrap Estimator Function

Bootstrapping requires defining a statistic function with signature `statistic(data, indices)`.

### Practice Problem
Write a function named `mean_fn` that:

1. Takes two arguments: `data` and `indices`.
2. Subsets the data vector using the indices: `d <- data[indices]`.
3. Returns the mean of the subsetted vector `d`.

```{r ex4, exercise=TRUE}
mean_fn <- function(data, indices) {
  # Write function here:
  
  
}
```

```{r ex4-hint-1}
# Hint: Define a function with arguments data and indices. Inside the function, subset the data vector using the indices, and return the mean of the subsetted data.
```

```{r ex4-solution}
  # solution/hint content purged from public repository history

```

```{r ex4-check}
grade_this_code()
```

---

## Exercise 5: Running the Bootstrap

The `boot` package runs the resampling algorithm and calculates standard errors.

### Practice Problem
Using the preloaded data frame `income$monthly_income` and custom `mean_fn`:

1. Use `boot()` to run `R = 500` bootstrap replications. Store the result in `boot_out`.
2. Extract the bootstrap standard error from `boot_out`. (Hint: Compute the standard deviation of bootstrap replicates `boot_out$t`).

```{r ex5, exercise=TRUE}
mean_fn <- function(data, indices) {
  mean(data[indices])
}

# 1. Run boot() for 500 replicates


# 2. Compute bootstrap standard error


```

```{r ex5-hint-1}
# Hint: Use boot() from the boot package specifying the data, your custom statistic function, and the number of replicates (R = 500). Calculate the standard deviation of the bootstrap replicates (stored in boot_out$t) to estimate the standard error.
```

```{r ex5-solution}
  # solution/hint content purged from public repository history

```

```{r ex5-check}
grade_this_code()
```

---

## Exercise 6: Bootstrap Percentile Confidence Intervals

The percentile interval extracts the alpha boundaries directly from sorted replicates.

### Practice Problem
Using the preloaded `boot_out` object:

1. Run `boot.ci()` to compute the 95% bootstrap confidence interval.
2. Select the `"perc"` (percentile) interval type explicitly.

```{r ex6, exercise=TRUE}
# Set up boot object
mean_fn <- function(data, indices) { mean(data[indices]) }
set.seed(123)
boot_out <- boot(data = income$monthly_income, statistic = mean_fn, R = 200)

# 1. Compute percentile bootstrap CI using boot.ci()


```

```{r ex6-hint-1}
# Hint: Use boot.ci() passing your boot object and setting type = 'perc'.
```

```{r ex6-solution}
  # solution/hint content purged from public repository history

```

```{r ex6-check}
grade_this_code()
```

---

## Exercise 7: Bootstrap BCa Confidence Intervals

The Bias-Corrected and Accelerated (BCa) interval adjusts for skewness and bias.

### Practice Problem
Using the preloaded `boot_out` object:

1. Run `boot.ci()` to compute the 95% BCa confidence interval.
2. Select the `"bca"` interval type explicitly.

```{r ex7, exercise=TRUE}
# Set up boot object
mean_fn <- function(data, indices) { mean(data[indices]) }
set.seed(123)
boot_out <- boot(data = income$monthly_income, statistic = mean_fn, R = 200)

# 1. Compute BCa bootstrap CI using boot.ci()


```

```{r ex7-hint-1}
# Hint: Use boot.ci() passing your boot object and setting type = 'bca'.
```

```{r ex7-solution}
  # solution/hint content purged from public repository history

```

```{r ex7-check}
grade_this_code()
```

---

## Exercise 8: Bootstrapping a Proportion

Bootstrapping a proportion requires resampling a binary indicator vector.

### Practice Problem
Consider a sample where 30 out of 100 subjects responded.

1. Create a binary indicator vector `resp <- c(rep(1, 30), rep(0, 70))`.
2. Define a function `prop_fn <- function(data, indices) { mean(data[indices]) }`.
3. Use `boot()` to run `R = 200` bootstrap replications of `resp` using `prop_fn`.

```{r ex8, exercise=TRUE}
# 1. Create binary vector resp


# 2. Define prop_fn


# 3. Run boot() on resp


```

```{r ex8-hint-1}
# Hint: Create a binary vector of ones and zeros. Define a proportion estimator function and pass it to boot() with R = 200.
```

```{r ex8-solution}
  # solution/hint content purged from public repository history

```

```{r ex8-check}
grade_this_code()
```

---

## Exercise 9: Bootstrapping a Correlation Coefficient

Bootstrapping can be used to estimate confidence intervals for correlation coefficients.

### Practice Problem
Let's bootstrap the correlation coefficient between variables `mpg` and `hp` in `mtcars`.

1. Write a function `cor_fn <- function(data, indices) { cor(data$mpg[indices], data$hp[indices]) }`.
2. Use `boot()` to run `R = 200` bootstrap replications on the `mtcars` data frame using `cor_fn`.

```{r ex9, exercise=TRUE}
# 1. Define cor_fn


# 2. Run boot() on mtcars data frame


```

```{r ex9-hint-1}
# Hint: Define a correlation estimator function that subsets both variables from the data frame based on indices. Pass this to boot() with the mtcars data frame.
```

```{r ex9-solution}
  # solution/hint content purged from public repository history

```

```{r ex9-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex9-quiz}
quiz(
  question("Which type of bootstrap confidence interval adjusts for both bias and skewness in the bootstrap distribution?",
    answer("Percentile interval", message = "No, percentile interval assumes a symmetric bootstrap distribution."),
    answer("BCa (Bias-Corrected and Accelerated) interval", correct = TRUE),
    incorrect = "Correct! BCa is preferred when the bootstrap distribution is skewed."
  )
)
```

**How to Report in a Research Paper:**

> A non-parametric bootstrap with 1,000 replications was used to estimate standard errors and confidence intervals. The BCa bootstrap 95% confidence interval was $BC_a [\$13,248, \$15,772]$.

---

## Concept Check Quiz

Test your understanding of Unit 3!

```{r quiz-u3-final}
quiz(
  caption = "Unit 3 Knowledge Check",
  question("How does increasing sample size n affect the width of a 95% Confidence Interval?",
    answer("It decreases the interval width (makes the estimate more precise).", correct = TRUE),
    answer("It increases the interval width.", message = "No, larger sample sizes reduce standard error."),
    answer("It has no effect on the interval width.", message = "Incorrect. Width is inversely proportional to sqrt(n)."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question_text(
    "What core R package function is used to generate bootstrap samples?",
    answer("boot", correct = TRUE),
    answer("boot()", correct = TRUE),
    allow_retry = TRUE,
    trim = TRUE,
    options = list(placeholder = "e.g., package::function_name")
  ),
  question_numeric(
    "What critical z-value corresponds to a standard 95% two-sided confidence interval? (Round to 2 decimal places)",
    answer(1.96, correct = TRUE),
    tolerance = 0.01,
    allow_retry = TRUE,
    correct = "Correct! z_0.025 = 1.96."
  ),
  question("Which bootstrap confidence interval method adjusts for both bias and skewness in the bootstrap distribution?",
    answer("Percentile Interval", message = "No, percentile intervals do not correct for acceleration/skewness."),
    answer("BCa (Bias-Corrected and Accelerated) Interval", correct = TRUE),
    answer("Standard Normal Interval", message = "No, normal intervals assume a symmetric normal sampling distribution."),
    allow_retry = TRUE
  ),
  question_numeric(
    "If sample size n = 30, what degrees of freedom (df) should be used for a one-sample t-test?",
    answer(29, correct = TRUE),
    allow_retry = TRUE,
    correct = "Correct! df = n - 1 = 29."
  )
)
```
