---
title: "Unit 2: Descriptive Statistics and Programming Fundamentals"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
scores <- read.csv("data/student_scores.csv")
```

## Welcome

Welcome to Unit 2! This tutorial covers descriptive statistics, programming structures (conditionals, loops, custom functions), S3 object-oriented programming, and modern R workflows (dplyr, ggplot2).

---


---

## Interactive Explorer: Skewness & Mean vs. Median

Observe how skewness affects central tendency in real time. Move the slider to shift data skewness and watch how the mean and median diverge!

```{r skew-demo-ui}
sliderInput("skew_param", "Distribution Skewness:", min = -2, max = 2, value = 0, step = 0.5)
plotOutput("skew_plot", height = "300px")
```

```{r skew-demo-server, context="server"}
output$skew_plot <- renderPlot({
  set.seed(123)
  n <- 500
  skew_val <- input$skew_param
  if (skew_val == 0) {
    x <- rnorm(n, mean = 50, sd = 10)
  } else if (skew_val > 0) {
    x <- rgamma(n, shape = 2 / skew_val, scale = 5 * skew_val) + 30
  } else {
    x <- 70 - rgamma(n, shape = 2 / abs(skew_val), scale = 5 * abs(skew_val))
  }
  
  m <- mean(x)
  med <- median(x)
  
  hist(x, breaks = 25, col = "#AED6F1", border = "white",
       main = paste("Mean =", round(m, 2), "| Median =", round(med, 2)),
       xlab = "Value")
  abline(v = m, col = "#C0392B", lwd = 3, lty = 1)
  abline(v = med, col = "#27AE60", lwd = 3, lty = 2)
  legend("topright", legend = c("Mean", "Median"), col = c("#C0392B", "#27AE60"), lwd = 3, lty = c(1, 2))
})
```


## Exercise 1: Central Tendency and Dispersion

Measures of central tendency and dispersion provide a numeric summary of data.

### Practice Problem
Using the preloaded data frame `scores` and base R functions:

1. Compute the mean score of students.
2. Compute the median score of students.
3. Compute the standard deviation and variance of scores.
4. Print the overall summary of scores using `summary()`.

```{r ex1, exercise=TRUE}
# 1. Calculate mean and median


# 2. Calculate standard deviation and variance


# 3. Print overall summary


```

```{r ex1-hint-1}
# Hint 1 (Concept): Access data frame columns using `df$column`. Calculate summary statistics using mean(), median(), sd(), var(), and summary().
```

```{r ex1-hint-2}
# Hint 2 (Code Skeleton):
mean(scores$score)
median(scores$score)
sd(scores$score)
var(scores$score)
summary(scores$score)
```

```{r ex1-hint-3}
# Hint 3 (Detailed Guide): `scores$score` selects the `score` column. Use standard base R summary functions.
```

```{r ex1-solution}
# 1. Calculate mean and median
mean(scores$score)
median(scores$score)

# 2. Calculate standard deviation and variance
sd(scores$score)
var(scores$score)

# 3. Print overall summary
summary(scores$score)
```

```{r ex1-check}
grade_this_code()
```

---

## Exercise 2: Custom Functions - Skewness

Custom functions allow you to implement algorithms not found in base R.

### Practice Problem
Recall the formula for sample skewness (biased denominator):
$$\text{Skewness} = \frac{\frac{1}{n}\sum(x_i - \bar{x})^3}{\left[\frac{1}{n}\sum(x_i - \bar{x})^2\right]^{3/2}}$$

Write a function named `skewness_manual` that:

1. Takes a numeric vector `x` as input.
2. Returns `NA` if `length(x) < 3`.
3. Computes and returns the skewness.

```{r ex2, exercise=TRUE}
skewness_manual <- function(x) {
  # Write your function logic here:
  
  
}

# Test your function:
test_vector <- c(4, 5, 6, 8, 10)
skewness_manual(test_vector)
```

```{r ex2-hint-1}
# Hint 1 (Concept): Create a function `skewness_manual <- function(x)`. Check if `length(x) < 3` and return NA. Calculate formula components using mean() and sum().
```

```{r ex2-hint-2}
# Hint 2 (Code Skeleton):
skewness_manual <- function(x) {
  n <- length(x)
  if (n < 3) return(NA)
  m <- mean(x)
  num <- sum((x - m)^3) / n
  den <- (sum((x - m)^2) / n)^(1.5)
  return(num / den)
}
```

```{r ex2-hint-3}
# Hint 3 (Detailed Guide): Standard skewness formula uses population variance in denominator: `(sum((x - m)^2)/n)^(3/2)`.
```

```{r ex2-solution}
skewness_manual <- function(x) {
  n <- length(x)
  if (n < 3) return(NA)
  m <- mean(x)
  num <- sum((x - m)^3) / n
  den <- (sum((x - m)^2) / n)^(1.5)
  return(num / den)
}

test_vector <- c(4, 5, 6, 8, 10)
skewness_manual(test_vector)
```

```{r ex2-check}
grade_this_code()
```

---

## Exercise 3: Custom Functions - Kurtosis

Kurtosis measures the "tailedness" of a theoretical probability distribution.

### Practice Problem
Recall the formula for sample kurtosis (biased denominator):
$$\text{Kurtosis} = \frac{\frac{1}{n}\sum(x_i - \bar{x})^4}{\left[\frac{1}{n}\sum(x_i - \bar{x})^2\right]^2}$$

Write a function named `kurtosis_manual` that:

1. Takes a numeric vector `x` as input.
2. Returns `NA` if `length(x) < 3`.
3. Computes and returns the kurtosis.

```{r ex3, exercise=TRUE}
kurtosis_manual <- function(x) {
  # Write your function logic here:
  
  
}

# Test your function:
test_vector <- c(4, 5, 6, 8, 10)
kurtosis_manual(test_vector)
```

```{r ex3-hint-1}
# Hint: Load the moments or psych package using library() or use the built-in function from those packages to calculate skewness and kurtosis.
```

```{r ex3-solution}
  # solution/hint content purged from public repository history

```

```{r ex3-check}
grade_this_code()
```

---

## Exercise 4: Conditional Statements

Conditional blocks allow branching logic based on tests.

### Practice Problem
Using an `if-else` block:

1. Write a block that checks a variable `age <- 20`.
2. If `age` is less than `18`, print `"Minor"`.
3. If `age` is between `18` and `65` (inclusive), print `"Adult"`.
4. Otherwise, print `"Senior"`.

```{r ex4, exercise=TRUE}
age <- 20

# Write your if-else conditional block here:


```

```{r ex4-hint-1}
# Hint: Initialize a counter or subset the vector. Use a loop (like for or while) to iterate through the indices, checking conditions using if/else statements. Increment your counts accordingly.
```

```{r ex4-solution}
  # solution/hint content purged from public repository history

```

```{r ex4-check}
grade_this_code()
```

---

## Exercise 5: Vectorized Conditionals

`ifelse()` is R's fast, vectorized conditional function.

### Practice Problem
Using the preloaded vector `scores_vec <- c(45, 78, 62, 85, 53)`:

1. Use `ifelse()` to create a vector named `grades`.
2. Assign `"Pass"` if a score is `60` or above, else assign `"Fail"`.

```{r ex5, exercise=TRUE}
scores_vec <- c(45, 78, 62, 85, 53)

# 1. Create grades vector using ifelse()


```

```{r ex5-hint-1}
# Hint: Use ifelse() for conditional element-wise operations on vectors, or logical indexing directly without looping.
```

```{r ex5-solution}
  # solution/hint content purged from public repository history

```

```{r ex5-check}
grade_this_code()
```

---

## Exercise 6: For Loops

For loops iterate over a predefined sequence.

### Practice Problem
1. Write a `for` loop that iterates through the numbers `1` to `5`.
2. If a number is even, print `"[number] is even"`, else print `"[number] is odd"` using `cat()` and the modulo operator `%%`.

```{r ex6, exercise=TRUE}
# Write your for loop here:


```

```{r ex6-hint-1}
# Hint: Define a function using function(arg) and return the output using return().
```

```{r ex6-solution}
  # solution/hint content purged from public repository history

```

```{r ex6-check}
grade_this_code()
```

---

## Exercise 7: While Loops

While loops execute as long as a logical condition remains true.

### Practice Problem
1. Initialize a counter `k <- 1`.
2. Write a `while` loop that runs as long as `k <= 3`.
3. Inside the loop, print `"Counter is:"` followed by `k`, and increment `k` by 1.

```{r ex7, exercise=TRUE}
# Write your while loop here:


```

```{r ex7-hint-1}
# Hint: Combine your custom function with loops or sapply() to calculate summaries across multiple columns.
```

```{r ex7-solution}
  # solution/hint content purged from public repository history

```

```{r ex7-check}
grade_this_code()
```

---

## Exercise 8: Repeat Loops with Break & Next

`repeat` creates an infinite loop that must be explicitly terminated with `break`.

### Practice Problem
1. Initialize `i <- 0`.
2. Write a `repeat` loop that increments `i` by 1 at each iteration.
3. If `i == 2`, skip printing by using `next`.
4. Print `i` using `print(i)`.
5. If `i >= 4`, terminate the loop using `break`.

```{r ex8, exercise=TRUE}
# Write your repeat loop here:


```

```{r ex8-hint-1}
# Hint: Use select() to subset specific columns by name or helper functions.
```

```{r ex8-solution}
  # solution/hint content purged from public repository history

```

```{r ex8-check}
grade_this_code()
```

---

## Exercise 9: Functions with Default Arguments

You can specify default values for function parameters in R.

### Practice Problem
1. Write a function named `power_func` that takes two arguments: `base` and `exponent`.
2. Set the default value of `exponent` to `2`.
3. The function should return `base^exponent`.
4. Test it by running `power_func(5)` and `power_func(5, 3)`.

```{r ex9, exercise=TRUE}
power_func <- function(base, exponent) {
  # Write function here:
  
  
}

# Test it:
power_func(5)
power_func(5, 3)
```

```{r ex9-hint-1}
# Hint: Use filter() to subset rows that match specific logical conditions (e.g. age > 30).
```

```{r ex9-solution}
  # solution/hint content purged from public repository history

```

```{r ex9-check}
grade_this_code()
```

---

## Exercise 10: S3 OOP - Creating S3 Objects

The S3 class system is R's informal, list-based OOP framework.

### Practice Problem
1. Create a list named `stud` containing elements: `name = "Alice"` and `score = 95`.
2. Set the class of `stud` to `"student"` using `class()`.
3. Return the class of `stud`.

```{r ex10, exercise=TRUE}
# 1. Create stud list


# 2. Assign class student


# 3. Return class


```

```{r ex10-hint-1}
# Hint: Use mutate() to create new columns or modify existing ones.
```

```{r ex10-solution}
  # solution/hint content purged from public repository history

```

```{r ex10-check}
grade_this_code()
```

---

## Exercise 11: S3 OOP - Writing Methods

S3 generic functions dispatch to methods using naming syntax: `generic.class`.

### Practice Problem
1. Create an S3 patient record list `pat <- list(name = "Bob", pulse = 75)`.
2. Set its class to `"patient"`.
3. Define the S3 method `print.patient <- function(x, ...)` to print `"Patient [name] pulse: [pulse]"` using `cat()`.

```{r ex11, exercise=TRUE}
# 1. Create pat patient object


# 2. Define print.patient method


# 3. Test printing pat


```

```{r ex11-hint-1}
# Hint: Combine group_by() and summarise() to calculate summary stats stratified by categories.
```

```{r ex11-solution}
  # solution/hint content purged from public repository history

```

```{r ex11-check}
grade_this_code()
```

---

## Exercise 12: Modern Data Manipulation with dplyr

`dplyr` provides high-performance data manipulation verbs.

### Practice Problem
Using the preloaded data frame `scores` and the pipe (`%>%` or `|>`):

1. **`filter`** for students in group `"A"`.
2. **`select`** only columns `group` and `score`.
3. **`mutate`** to create `score_percent` dividing `score` by 100.
4. **`group_by`** group and **`summarise`** to find the mean and standard deviation of score.

```{r ex12, exercise=TRUE}
# 1. Filter, select, and mutate in one pipeline


# 2. Group by and calculate group means


```

```{r ex12-hint-1}
# Hint: Use ggplot() specifying the data frame and aes(x, y). Add points using geom_point().
```

```{r ex12-solution}
  # solution/hint content purged from public repository history

```

```{r ex12-check}
grade_this_code()
```

---

## Exercise 13: Data Visualization with ggplot2

`ggplot2` builds visual plots incrementally using layers.

### Practice Problem
Using the preloaded data frame `scores`:

1. Create a scatter plot of `hours_studied` (x-axis) vs `score` (y-axis) using `geom_point()`.
2. Map `group` to the color aesthetic.
3. Add a linear regression line using `geom_smooth(method = "lm", se = FALSE)`.

```{r ex13, exercise=TRUE}
# Create a scatter plot of hours_studied vs score with group color and a linear trend line


```

```{r ex13-hint-1}
# Hint: Use ggplot() specifying the data frame and aes(x, y). Add box plots using geom_boxplot() and color them by group.
```

```{r ex13-solution}
  # solution/hint content purged from public repository history

```

```{r ex13-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex13-quiz}
quiz(
  question("What is the S3 method naming convention in R?",
    answer("`generic_name.class_name`", correct = TRUE),
    answer("`class_name$generic_name`", message = "No, that represents class attributes or environments."),
    incorrect = "Correct! For print on patient, it is print.patient."
  )
)
```

**How to Report in a Research Paper:**

> S3 object-oriented systems were used in R to create structured records, facilitating automated custom formatting.

---

## Concept Check Quiz

Test your understanding of Unit 2!

```{r quiz}
quiz(
  question("Which of the following dplyr verbs is used to subset columns of a data frame?",
    answer("`filter()`", message = "No, filter() subsets rows."),
    answer("`select()`", correct = TRUE),
    answer("`mutate()`", message = "No, mutate() adds or modifies columns."),
    incorrect = "Correct! select() subsets variables (columns)."
  ),
  question("Which loop construct is guaranteed to run indefinitely if the stopping condition is never updated inside the loop?",
    answer("`for` loop", message = "No, for loops terminate once the sequence ends."),
    answer("`while` loop", correct = TRUE),
    incorrect = "Correct! A while loop runs as long as its condition remains TRUE."
  )
)
```