---
title: "Unit 5: Non-Parametric Tests and Applications"
description: "Hands-on tutorial on Wilcoxon, Mann-Whitney U, Kruskal-Wallis, and Friedman non-parametric tests in R."
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE)
course_df <- read.csv("data/course_dataset.csv")
exam_df <- read.csv("data/exam_pairs.csv")
```

## Welcome

Welcome to Unit 5! This tutorial covers non-parametric statistical methods, including the exact binomial test, sign test, Wilcoxon tests (rank-sum and signed-rank), Kruskal-Wallis, Friedman, Kendall's W, post-hoc procedures, and clinical research reporting.

---


---

## Interactive Explorer: Outlier Robustness (Mann-Whitney vs. t-Test)

Drag the slider to introduce a severe outlier into Group 1 and observe how the $t$-test $p$-value breaks down, while the Mann-Whitney $U$ test stays robust!

```{r outlier-demo-ui}
sliderInput("outlier_val", "Outlier Value added to Group 1:", min = 0, max = 200, value = 0, step = 10)
plotOutput("outlier_plot", height = "300px")
```

```{r outlier-demo-server, context="server"}
output$outlier_plot <- renderPlot({
  set.seed(101)
  g1 <- c(rnorm(14, mean = 20, sd = 3), 20 + input$outlier_val)
  g2 <- rnorm(15, mean = 25, sd = 3)
  
  t_res <- t.test(g1, g2)
  mw_res <- wilcox.test(g1, g2)
  
  df_sim <- data.frame(
    val = c(g1, g2),
    grp = rep(c("Group 1 (with outlier)", "Group 2"), times = c(15, 15))
  )
  
  boxplot(val ~ grp, data = df_sim, col = c("#E74C3C", "#2ECC71"),
          main = paste("t-test p =", round(t_res$p.value, 4), "| Mann-Whitney p =", round(mw_res$p.value, 4)),
          ylab = "Value")
})
```


## Exercise 1: Exact Binomial Test

The exact binomial test evaluates whether observed proportions deviate from a hypothesized probability.

### Practice Problem
Let's test if the proportion of patients receiving the `"New"` treatment differs significantly from a hypothesized 50% (`p = 0.5`).

1. Count the number of patients receiving the `"New"` treatment in `course_df$treatment` and store it in `n_new`.
2. Compute the total number of patients and store it in `n_total`.
3. Use `binom.test()` to perform the exact binomial test.

```{r ex1, exercise=TRUE}
# 1. Count patients in New treatment group


# 2. Count total patients


# 3. Run binom.test()


```

```{r ex1-hint-1}
# Hint 1 (Concept): Count successes in the treatment vector using sum(course_df$treatment == 'New'), get total length using nrow(course_df), and call binom.test(n_new, n_total, p = 0.5).
```

```{r ex1-hint-2}
# Hint 2 (Code Skeleton):
n_new <- sum(course_df$treatment == "New")
n_total <- nrow(course_df)
binom.test(n_new, n_total, p = 0.5)
```

```{r ex1-hint-3}
# Hint 3 (Detailed Guide): `binom.test(x, n, p)` evaluates if observed count `x` out of `n` trials differs from theoretical probability `p`.
```

```{r ex1-solution}
# 1. Count patients in New treatment group
n_new <- sum(course_df$treatment == "New")

# 2. Count total patients
n_total <- nrow(course_df)

# 3. Run binom.test()
binom.test(n_new, n_total, p = 0.5)
```

```{r ex1-check}
grade_this_code()
```

---

## Exercise 2: Sign Test

The sign test evaluates whether the median of a variable differs from a hypothesized value.

### Practice Problem
Let's test if the median week 8 pain score (`week8_pain` in `course_df`) differs from a hypothesized clinical threshold of `4.0`.

1. Compute the differences `diffs` between each score and `4.0`.
2. Count the number of positive differences `n_pos` (greater than 0).
3. Determine the number of valid (non-tied) cases `n_valid` (excluding exactly 0 differences).
4. Use `binom.test()` to perform the two-tailed sign test.

```{r ex2, exercise=TRUE}
# 1. Compute differences from threshold 4.0


# 2. Count positive differences n_pos


# 3. Count non-tied differences n_valid


# 4. Perform binom.test()


```

```{r ex2-hint-1}
# Hint 1 (Concept): Calculate diffs = course_df$week8_pain - 4.0. Count n_pos = sum(diffs > 0) and n_valid = sum(diffs != 0). Run binom.test(n_pos, n_valid, p = 0.5).
```

```{r ex2-hint-2}
# Hint 2 (Code Skeleton):
diffs <- course_df$week8_pain - 4.0
n_pos <- sum(diffs > 0)
n_valid <- sum(diffs != 0)
binom.test(n_pos, n_valid, p = 0.5)
```

```{r ex2-hint-3}
# Hint 3 (Detailed Guide): Ties (where difference is exactly 0) must be discarded when performing a Sign Test. `binom.test(n_pos, n_valid, p = 0.5)` tests if positive differences exceed 50%.
```

```{r ex2-solution}
# 1. Compute differences from threshold 4.0
diffs <- course_df$week8_pain - 4.0

# 2. Count positive differences n_pos
n_pos <- sum(diffs > 0)

# 3. Count non-tied differences n_valid
n_valid <- sum(diffs != 0)

# 4. Perform binom.test()
binom.test(n_pos, n_valid, p = 0.5)
```

```{r ex2-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex2-quiz}
quiz(
  question("Based on the binom.test result, is the median week 8 pain score significantly different from the clinical threshold of 4.0?",
    answer("Yes, the p-value is extremely small (p < 0.001), indicating a statistically significant difference.", correct = TRUE),
    answer("No, the p-value is greater than 0.05.", message = "Incorrect. Check the p-value returned by the binomial test."),
    incorrect = "Correct! The p-value is way below 0.05, so the difference is highly significant."
  )
)
```

**How to Report in a Research Paper:**

> A sign test was conducted to determine whether the median week 8 pain score differed from a clinical threshold of 4.0. The median pain score (3.0) was significantly lower than the hypothesized threshold, $p < 0.001$.

---

## Exercise 3: Wilcoxon Rank-Sum Test (Mann-Whitney U)

The Wilcoxon Rank-Sum test compares distributions of two independent groups.

### Practice Problem
Using `course_df`:

1. Run a Wilcoxon Rank-Sum test to compare `week8_pain` levels between treatment groups (`treatment`). Use formula notation: `week8_pain ~ treatment`.

```{r ex3, exercise=TRUE}
# 1. Run Wilcoxon Rank-Sum test


```

```{r ex3-hint-1}
# Hint: Use wilcox.test() with formula notation (variable ~ group) on the dataset.
```

```{r ex3-solution}
  # solution/hint content purged from public repository history

```

```{r ex3-check}
grade_this_code()
```

---

## Exercise 4: Wilcoxon Signed-Rank Test (Paired)

The Wilcoxon Signed-Rank test evaluates differences in paired (repeated measures) observations.

### Practice Problem
Using the preloaded data frame `exam_df`:

1. Perform a paired Wilcoxon Signed-Rank test to evaluate whether student scores increased from `pre_test` to `post_test`.
2. Remember to enable `paired = TRUE`.

```{r ex4, exercise=TRUE}
# 1. Run paired Wilcoxon test on exam_df


```

```{r ex4-hint-1}
# Hint: Use wilcox.test() specifying the pre-test and post-test vectors, and enable paired = TRUE.
```

```{r ex4-solution}
  # solution/hint content purged from public repository history

```

```{r ex4-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex4-quiz}
quiz(
  question("What does the Wilcoxon Signed-Rank test (paired) show about the change in scores from pre_test to post_test?",
    answer("Scores significantly increased from pre_test to post_test (p < 0.01).", correct = TRUE),
    answer("There was no statistically significant change in scores (p > 0.05).", message = "Incorrect. Check the paired Wilcoxon test p-value."),
    incorrect = "Correct! The paired Wilcoxon signed-rank test yields a p-value less than 0.05, indicating a significant improvement."
  )
)
```

**How to Report in a Research Paper:**

> A Wilcoxon Signed-Rank test was performed to evaluate the change in student exam scores. The results indicated a statistically significant increase from pre-test to post-test scores, $V = 8.5, p < 0.01$.

---

## Exercise 5: Kruskal-Wallis Test

The Kruskal-Wallis test compares medians across three or more independent groups.

### Practice Problem
Let's test if the week 8 pain levels (`week8_pain`) differ across the three hospital sites (`site` in `course_df`).

1. Run the overall `kruskal.test()` with the formula `week8_pain ~ site` and dataset `course_df`.

```{r ex5, exercise=TRUE}
# 1. Run overall Kruskal-Wallis test


```

```{r ex5-hint-1}
# Hint: Use kruskal.test() with formula notation (variable ~ group) on the dataset.
```

```{r ex5-solution}
  # solution/hint content purged from public repository history

```

```{r ex5-check}
grade_this_code()
```

---

## Exercise 6: Post-Hoc Pairwise Wilcoxon Tests

Post-hoc pairwise Wilcoxon tests identify which specific groups differ after a significant Kruskal-Wallis result.

### Practice Problem
Using `course_df`:

1. Run post-hoc pairwise Wilcoxon Rank-Sum tests using `pairwise.wilcox.test()`.
2. Apply a Bonferroni p-value adjustment.

```{r ex6, exercise=TRUE}
# 1. Run pairwise Wilcoxon tests with Bonferroni correction


```

```{r ex6-hint-1}
# Hint: Use pairwise.wilcox.test() passing the numeric variable and grouping factor, specifying p.adjust.method = 'bonferroni'.
```

```{r ex6-solution}
  # solution/hint content purged from public repository history

```

```{r ex6-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex6-quiz}
quiz(
  question("Based on the overall Kruskal-Wallis test, do pain scores differ significantly across the three hospital sites?",
    answer("No, there is no significant difference (p > 0.05).", correct = TRUE),
    answer("Yes, there is a statistically significant difference across sites (p < 0.001).", message = "Incorrect. The Kruskal-Wallis test p-value is 0.455, which is greater than 0.05."),
    incorrect = "Correct! The Kruskal-Wallis test is non-significant, showing no systematic differences across sites."
  )
)
```

**How to Report in a Research Paper:**

> A Kruskal-Wallis test showed no significant difference in week 8 pain scores across the three hospital sites, $\chi^2(2) = 1.57, p = 0.455$.

---

## Exercise 7: Friedman Test

The Friedman test evaluates differences in repeated measures designs across three or more conditions.

### Practice Problem
Let's analyze pain scores at three time points (baseline, week 4, week 8) structured inside `course_df` as columns `baseline_pain`, `week4_pain`, and `week8_pain`.

1. The Friedman test requires data in a long format. Reshape the columns into a long format using the pre-built code below.
2. Run `friedman.test()` using the long format formula `pain ~ time | id`.

```{r ex7, exercise=TRUE}
# Set up long format data
n_subjects <- nrow(course_df)
long_pain <- data.frame(
  id = rep(1:n_subjects, times = 3),
  time = rep(c("baseline", "week4", "week8"), each = n_subjects),
  pain = c(course_df$baseline_pain, course_df$week4_pain, course_df$week8_pain)
)

# 1. Run the Friedman test on long_pain


```

```{r ex7-hint-1}
# Hint: Use friedman.test() with formula notation (variable ~ time | subject_id) on the reshaped long dataset.
```

```{r ex7-solution}
  # solution/hint content purged from public repository history

```

```{r ex7-check}
grade_this_code()
```

---

## Exercise 8: Kendall's W Effect Size

Kendall's W measures the effect size of repeated measures Friedman ANOVA.

### Practice Problem
Compute **Kendall's W** effect size from the Friedman Chi-Squared statistic ($Q$):
$$W = \frac{Q}{N(k-1)}$$
Where $Q = 264.98$ is the Friedman test statistic, $N = 180$ is the sample size, and $k = 3$ is the number of time points.

```{r ex8, exercise=TRUE}
Q <- 264.98
N <- 180
k <- 3

# 1. Compute Kendall's W


```

```{r ex8-hint-1}
# Hint: Compute Kendall's W manually using the Friedman chi-squared statistic, sample size, and number of conditions.
```

```{r ex8-solution}
  # solution/hint content purged from public repository history

```

```{r ex8-check}
grade_this_code()
```

---

## Exercise 9: Non-Parametric Test Selection Scenarios

Selecting the appropriate statistical test is essential for robust data analysis.

### Practice Problem
Assume we are testing student performance grades (ordinal scales) across 4 different class sections (independent groups). Which test is appropriate?
Write the name of the correct R function as a string (either `"wilcox.test"` or `"kruskal.test"` or `"friedman.test"`).

```{r ex9, exercise=TRUE}
# Write the correct function name as a string:


```

```{r ex9-hint-1}
# Hint: Recall the appropriate non-parametric analog for comparing ordinal scores across three or more independent groups.
```

```{r ex9-solution}
  # solution/hint content purged from public repository history

```

```{r ex9-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex9-quiz}
quiz(
  question("What does the Friedman test and its effect size suggest about pain levels over time?",
    answer("Pain levels changed significantly over the three time points (p < 0.001) with a moderate-to-strong effect size (Kendall's W = 0.74).", correct = TRUE),
    answer("There was a significant change but with a negligible effect size.", message = "Incorrect. Kendall's W is 0.74, which indicates a strong effect."),
    incorrect = "Correct! The Friedman test shows highly significant differences across time points, and Kendall's W indicates a strong effect size."
  )
)
```

**How to Report in a Research Paper:**

> A Friedman test was conducted to compare pain scores across three time points (baseline, week 4, week 8). Pain levels differed significantly across time, $\chi^2(2) = 264.98, p < 0.001$, with a strong effect size, Kendall's $W = 0.74$.

---

## Concept Check Quiz

Test your understanding of Unit 5!

```{r quiz-u5-final}
quiz(
  caption = "Unit 5 Knowledge Check (10 Questions)",
  question("Why are non-parametric tests like Mann-Whitney U resistant to extreme outliers?",
    answer("Because they operate on data ranks rather than raw numerical values.", correct = TRUE),
    answer("Because they automatically delete outliers before testing.", message = "No, non-parametric tests preserve all data points by converting them to ranks."),
    answer("Because they assume a normal distribution.", message = "Incorrect. Non-parametric tests make no distributional assumptions."),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question_text(
    "What R function is used to conduct a non-parametric one-way ANOVA across 3+ independent groups?",
    answer("kruskal.test", correct = TRUE), answer("kruskal.test()", correct = TRUE),
    allow_retry = TRUE, trim = TRUE, options = list(placeholder = "e.g., function_name")
  ),
  question("Which non-parametric test is used for repeated measures / blocked designs across 3+ related groups?",
    answer("Friedman test", correct = TRUE),
    answer("Mann-Whitney U test", message = "No, Mann-Whitney U is for 2 independent groups."),
    answer("Wilcoxon rank-sum test", message = "No, Wilcoxon rank-sum is for 2 independent groups."),
    allow_retry = TRUE
  ),
  question("What is the non-parametric equivalent of the paired Student's t-test for 2 related samples?",
    answer("Wilcoxon Signed-Rank Test", correct = TRUE),
    answer("Mann-Whitney U Test", message = "No, Mann-Whitney U is for independent samples."),
    allow_retry = TRUE
  ),
  question_numeric(
    "What is the minimum number of independent treatment groups required for a Kruskal-Wallis test?",
    answer(3, correct = TRUE), allow_retry = TRUE,
    correct = "Correct! Kruskal-Wallis extends Mann-Whitney to k >= 3 groups."
  ),
  question("Which R function conducts an Exact Binomial Test for a single proportion?",
    answer("`binom.test()`", correct = TRUE),
    answer("`prop.test()`", message = "No, prop.test() uses normal approximation."),
    allow_retry = TRUE
  ),
  question("Which non-parametric test is used to compare the empirical cumulative distribution function (ECDF) of two samples?",
    answer("Kolmogorov-Smirnov test (`ks.test()`)", correct = TRUE),
    answer("Wilcoxon test", message = "No, Wilcoxon compares location/ranks."),
    allow_retry = TRUE
  ),
  question("When normal distribution assumptions ARE met, what is the relative asymptotic efficiency of the Mann-Whitney U test compared to Student's t-test?",
    answer("Approximately 95.5%", correct = TRUE),
    answer("Approximately 50%", message = "No, non-parametric tests lose very little power (~95.5%)."),
    allow_retry = TRUE
  ),
  question("How does wilcox.test() handle ties (identical values) when exact = TRUE?",
    answer("It emits a warning and falls back to normal asymptotic approximation.", correct = TRUE),
    answer("It throws a fatal error.", message = "No, it falls back to normal approximation."),
    allow_retry = TRUE
  ),
  question("Which test compares paired binary responses (e.g. success/failure before vs after treatment)?",
    answer("`mcnemar.test()`", correct = TRUE),
    answer("`chisq.test()`", message = "No, chisq.test() is for independent samples."),
    allow_retry = TRUE
  )
)
```
