---
title: "Unit 4: Analysis of Contingency Tables"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
library(tableone)
library(gtsummary)
knitr::opts_chunk$set(echo = FALSE)
scores <- read.csv("data/student_scores.csv")
```

## Welcome

Welcome to Unit 4! This tutorial covers generating tables, Chi-Square tests (with/without Yates' correction), calculating odds ratios, Cohen's Kappa, trend tests, paired designs, stratified analysis, and compact baseline and summary tables.

---

## Exercise 1: Creating Contingency Tables

A contingency table displays the frequency distribution of categorical variables.

### Practice Problem
Consider a clinical trial dataset with 180 subjects. We have two categorical vectors: `treatment` (New vs. Standard) and `pain_relief` (Relieved vs. Not Relieved).

1. Create a 2x2 contingency table named `tab` using the `table()` function.
2. Add marginal row and column totals using `addmargins()`.

```{r ex1, exercise=TRUE}
treatment <- rep(c("New", "Standard"), each = 90)
pain_relief <- c(rep(c("Relieved", "Not Relieved"), times = c(75, 15)),
                 rep(c("Relieved", "Not Relieved"), times = c(30, 60)))

# 1. Create contingency table tab


# 2. Add margins to tab


```

```{r ex1-hint-1}
# Hint 1 (Concept): Use table(treatment, pain_relief) to create the 2x2 table, then pass it to addmargins() to add row/col totals.
```

```{r ex1-hint-2}
# Hint 2 (Code Skeleton):
tab <- table(treatment, pain_relief)
addmargins(tab)
```

```{r ex1-hint-3}
# Hint 3 (Detailed Guide): `table()` summarizes frequencies across the two categorical factors. `addmargins()` appends Sum columns and rows.
```

```{r ex1-hint-2}
  # solution/hint content purged from public repository history

```

```{r ex1-solution}
treatment <- rep(c("New", "Standard"), each = 90)
pain_relief <- c(rep(c("Relieved", "Not Relieved"), times = c(75, 15)),
                 rep(c("Relieved", "Not Relieved"), times = c(30, 60)))

# 1. Create contingency table tab
tab <- table(treatment, pain_relief)

# 2. Add margins to tab
addmargins(tab)
```

```{r ex1-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex1-quiz}
quiz(
  question("What is the total number of subjects in the contingency table (including row and column margins)?",
    answer("180 subjects", correct = TRUE),
    answer("90 subjects", message = "Incorrect. While each treatment group has 90 subjects, the total sample size is 180."),
    incorrect = "Correct! The sum of all cell counts is 180 subjects."
  )
)
```

**How to Report in a Research Paper:**

> Subjects were cross-classified by treatment (New vs. Standard) and pain relief (Relieved vs. Not Relieved) to form a 2x2 contingency table ($N = 180$).

---

## Exercise 2: Yates-Corrected Chi-Square Test

Pearson's Chi-Square test determines whether there is a statistically significant association between two categorical variables.

### Practice Problem
Using the preloaded table `tab` representing Treatment vs Relief:

1. Run a Pearson's Chi-Square test with Yates' continuity correction (R's default behavior).

```{r ex2, exercise=TRUE}
tab <- matrix(c(15, 75, 60, 30), nrow = 2, byrow = TRUE,
              dimnames = list(Treatment = c("New", "Standard"),
                              Relief = c("Not Relieved", "Relieved")))

# 1. Run Yates-corrected Chi-Square


```

```{r ex2-hint-1}
# Hint 1 (Concept): Pass the 2x2 matrix or table to chisq.test(). By default, correct = TRUE applies Yates' continuity correction.
```

```{r ex2-hint-2}
# Hint 2 (Code Skeleton):
chisq.test(tab)
```

```{r ex2-hint-3}
# Hint 3 (Detailed Guide): `chisq.test(tab)` performs Pearson's Chi-Square test with Yates' continuity correction for 2x2 tables.
```

```{r ex2-solution}
tab <- matrix(c(15, 75, 60, 30), nrow = 2, byrow = TRUE,
              dimnames = list(Treatment = c("New", "Standard"),
                              Relief = c("Not Relieved", "Relieved")))

# 1. Run Yates-corrected Chi-Square
chisq.test(tab)
```

```{r ex2-check}
grade_this_code()
```

---

## Exercise 3: Chi-Square Test Without Correction

When sample sizes are sufficiently large, continuity correction is not necessary.

### Practice Problem
Using the same table `tab`:

1. Run a standard Pearson's Chi-Square test *without* continuity correction (disabling it).

```{r ex3, exercise=TRUE}
tab <- matrix(c(15, 75, 60, 30), nrow = 2, byrow = TRUE,
              dimnames = list(Treatment = c("New", "Standard"),
                              Relief = c("Not Relieved", "Relieved")))

# 1. Run Chi-Square without correction


```

```{r ex3-hint-1}
# Hint: Use chisq.test() on the contingency table, setting correct = FALSE.
```

```{r ex3-solution}
  # solution/hint content purged from public repository history

```

```{r ex3-check}
grade_this_code()
```

---

## Exercise 4: Odds Ratio Calculation

The Odds Ratio (OR) quantifies the strength of association between two events.

### Practice Problem
Using the cell counts in the table `tab`:

1. Compute the **Odds Ratio (OR)** of achieving relief (New vs. Standard) using cell values manually:
   $$OR = \frac{a \times d}{b \times c}$$

```{r ex4, exercise=TRUE}
tab <- matrix(c(15, 75, 60, 30), nrow = 2, byrow = TRUE,
              dimnames = list(Treatment = c("New", "Standard"),
                              Relief = c("Not Relieved", "Relieved")))

# 1. Calculate Odds Ratio manually


```

```{r ex4-hint-1}
# Hint: Calculate the odds of success for each group manually, then divide them to obtain the odds ratio.
```

```{r ex4-solution}
  # solution/hint content purged from public repository history

```

```{r ex4-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex4-quiz}
quiz(
  question("Based on the Yates-corrected Chi-Square test result, is the association between treatment and pain relief statistically significant?",
    answer("Yes, the p-value is extremely small (p < 0.001), indicating a highly significant association.", correct = TRUE),
    answer("No, the p-value is greater than 0.05.", message = "Incorrect. The p-value returned by the Chi-Square test is way below 0.05."),
    incorrect = "Correct! The association between treatment type and pain relief outcome is highly significant."
  )
)
```

**How to Report in a Research Paper:**

> A Yates-corrected Chi-Square test of independence revealed a significant association between treatment type and pain relief, $\chi^2(1) = 43.2, p < 0.001$. Subjects receiving the New treatment had significantly higher odds of achieving pain relief than those receiving the Standard treatment, Odds Ratio (OR) = 10.0.

---

## Exercise 5: Cohen's Kappa

Cohen's Kappa measures inter-rater agreement for categorical items.

### Practice Problem
Two clinical observers rated 50 X-ray scans as "Normal" or "Abnormal".

- Observer A & B both Normal: 30
- Observer A Normal, Observer B Abnormal: 5
- Observer A Abnormal, Observer B Normal: 3
- Observer A & B both Abnormal: 12

1. Compute observed agreement ($p_o$) and expected agreement by chance ($p_e$).
2. Calculate Cohen's Kappa ($\kappa$):
   $$\kappa = \frac{p_o - p_e}{1 - p_e}$$

```{r ex5, exercise=TRUE}
kappa_tab <- matrix(c(30, 5, 3, 12), nrow = 2, byrow = TRUE)

# 1. Compute po and pe


# 2. Calculate Cohen's Kappa


```

```{r ex5-hint-1}
# Hint: Calculate observed agreement (diagonal sum divided by total) and expected agreement by chance based on marginal totals. Compute Kappa using the standard formula.
```

```{r ex5-solution}
  # solution/hint content purged from public repository history

```

```{r ex5-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex5-quiz}
quiz(
  question("What is the calculated value of Cohen's Kappa, and how is it interpreted?",
    answer("Kappa = 0.63, indicating substantial agreement between observers.", correct = TRUE),
    answer("Kappa = 0.15, indicating slight agreement.", message = "Incorrect. Check your Kappa calculation output."),
    incorrect = "Correct! According to Landis & Koch criteria, Kappa = 0.63 indicates substantial agreement."
  )
)
```

**How to Report in a Research Paper:**

> Inter-rater reliability between the two clinical observers rating X-ray scans showed substantial agreement, Cohen's $\kappa = 0.63$.

---

## Exercise 6: Cochran-Armitage Trend Test

The trend test evaluates linear trends in binomial proportions across ordered categories.

### Practice Problem
Perform a Cochran-Armitage Trend Test using `prop.trend.test(x, n)` where successes are `x = c(10, 20, 35)` and sample sizes are `n = c(50, 50, 50)`.

```{r ex6, exercise=TRUE}
# 1. Run Cochran-Armitage Trend Test


```

```{r ex6-hint-1}
# Hint: Use prop.trend.test() with the successes vector and total sample size vector.
```

```{r ex6-solution}
  # solution/hint content purged from public repository history

```

```{r ex6-check}
grade_this_code()
```

---

## Exercise 7: McNemar's Paired Test

McNemar's test is used on paired nominal data (pre vs post).

### Practice Problem
Perform a McNemar test on a paired matrix `paired_tab` (pre vs post results for same subjects).

```{r ex7, exercise=TRUE}
# Paired table matrix:
paired_tab <- matrix(c(40, 10, 25, 15), nrow = 2)

# 1. Run McNemar Paired Test on paired_tab


```

```{r ex7-hint-1}
# Hint: Use mcnemar.test() on the paired contingency table matrix.
```

```{r ex7-solution}
  # solution/hint content purged from public repository history

```

```{r ex7-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex7-quiz}
quiz(
  question("Based on the McNemar Paired Test result, does the proportion of pain relief differ significantly before and after treatment?",
    answer("Yes, McNemar's test is statistically significant (p = 0.018 < 0.05), indicating a significant change.", correct = TRUE),
    answer("No, the p-value is greater than 0.05.", message = "Incorrect. Check your McNemar test p-value output."),
    incorrect = "Correct! The paired proportion difference is statistically significant."
  )
)
```

**How to Report in a Research Paper:**

> A McNemar test was used to analyze paired pre- and post-treatment outcomes. The proportion of subjects reporting pain relief differed significantly between pre- and post-treatment phases, $\chi^2(1) = 5.8, p = 0.018$.

---

## Exercise 8: Cochran-Mantel-Haenszel Test

The CMH test evaluates the association between binary variables while controlling for confounding strata.

### Practice Problem
Perform a Cochran-Mantel-Haenszel (CMH) test on a 2x2x2 stratified array `strat_array`.

```{r ex8, exercise=TRUE}
# Stratified 2x2x2 array (Stratum 1 & 2):
strat_array <- array(c(10, 20, 15, 30,  # Stratum 1
                       5, 15, 8, 22),   # Stratum 2
                     dim = c(2, 2, 2))

# 1. Run Cochran-Mantel-Haenszel Test on strat_array


```

```{r ex8-hint-1}
# Hint: Use mantelhaen.test() on the 3D stratified contingency table array.
```

```{r ex8-solution}
  # solution/hint content purged from public repository history

```

```{r ex8-check}
grade_this_code()
```

---

## Exercise 9: Baseline Tables using tableone

The `tableone` package simplifies the creation of compact baseline characteristic tables.

### Practice Problem
Using the preloaded data frame `scores`:

1. Create a baseline summary table stratified by `group`.
2. Use variables `score` and `hours_studied` as variables.

```{r ex9, exercise=TRUE}
# 1. Create tableone stratified by group


```

```{r ex9-hint-1}
# Hint: Use CreateTableOne() specifying the variables list, strata category, and the data frame.
```

```{r ex9-solution}
  # solution/hint content purged from public repository history

```

```{r ex9-check}
grade_this_code()
```

---

## Exercise 10: Summary Tables using gtsummary

The `gtsummary` package creates presentation-ready summary tables for publications.

### Practice Problem
Using the preloaded data frame `scores`:

1. Use `tbl_summary()` to create a summary table of scores stratified by `group`.

```{r ex10, exercise=TRUE}
# 1. Create gtsummary table


```

```{r ex10-hint-1}
# Hint: Use tbl_summary() on the data frame specifying the strata category using the by argument.
```

```{r ex10-solution}
  # solution/hint content purged from public repository history

```

```{r ex10-check}
grade_this_code()
```

### Check Your Understanding & How to Report

```{r ex10-quiz}
quiz(
  question("Under what condition is Fisher's Exact Test preferred over Pearson's Chi-Square Test?",
    answer("When any expected cell count in the contingency table is less than 5.", correct = TRUE),
    answer("When the sample size is extremely large.", message = "Incorrect. Chi-Square is suitable for large samples; Fisher's is exact and preferred for small cell counts."),
    incorrect = "Correct! Fisher's Exact Test is preferred when expected cell frequencies are small (e.g. less than 5)."
  )
)
```

**How to Report in a Research Paper:**

> Demographics and baseline clinical characteristics were summarized by treatment group using the tableone and gtsummary packages in R.

---

## Concept Check Quiz

Test your understanding of Unit 4!

```{r quiz}
quiz(
  question("Which test is appropriate for evaluating differences in categorical outcomes from paired observations (e.g. before/after on the same patients)?",
    answer("Pearson's Chi-Square test", message = "No, standard Chi-Square assumes independent observations."),
    answer("McNemar's test", correct = TRUE),
    answer("Cochran-Mantel-Haenszel test", message = "No, CMH is for stratified independent tables."),
    incorrect = "McNemar's test checks marginal homogeneity in paired contingency tables."
  )
)
```